---
title: "Data Science Lab 7 : Making Maps"
author: "Lauren"
date: "5/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Goals of this lab

* Practice creating static and interactive choropleth maps.
* Practice creating static maps with `ggmap` and interactive maps with `leaflet`.


```{r}
# Load libraries
library(tidyverse)
library(ggmap)
library(leaflet)
library(RColorBrewer)
library(tidycensus)
library(gganimate)
library(pdxTrees)
library(here)
```



### Problem 1: Mapping PDX Crashes

For this problem we will return to the SE Portland 2018 car crash dataset. (( ODOT Crash records for SE PDX, all road types, for Jan 1, 2018 - Dec 21, 2018. Downloaded from the ODOT Crash Data System 11/11/2019. Accessed from [ODOT](https://www.oregon.gov/ODOT/Data/Pages/Crash.aspx) ))

```{r}
pdx_crash_2018 <- read_csv(here("Maps/pdx_crash_2018_page1.csv"))
```


d. Create a factor variable, `day_time` with categories:

* morning: crashes between 5am and noon
* afternoon: crashes after noon but before or at 4pm
* evening: crashes after 4pm but before or at 8pm
* night: crashes after 8pm but before or at 5am

Add this variable to your interactive map using color.  Make sure to include a legend and be mindful of your color palette choice.

```{r}
pdx_crash_2018 <- pdx_crash_2018 %>%
  mutate(CRASH_HR_NO = as.numeric(CRASH_HR_NO),
         day_time = case_when(
            CRASH_HR_NO > 5 & CRASH_HR_NO < 12 ~ "morning",
            CRASH_HR_NO >= 12 & CRASH_HR_NO <= 16 ~ "afternoon",
            CRASH_HR_NO > 16 & CRASH_HR_NO <= 20 ~ "evening",
            CRASH_HR_NO > 20 & CRASH_HR_NO <= 23 ~ "night",
            CRASH_HR_NO <= 5 ~ "night"),
         day_time = as.factor(day_time),
         day_time = fct_relevel(day_time, c("morning", "afternoon",
                                            "evening", "night")))

levels(pdx_crash_2018$day_time)

pal <- colorFactor(palette = "Dark2", 
                   domain = pdx_crash_2018$day_time)

leaflet(options = leafletOptions(minZoom = 11, maxZoom = 16)) %>% 
  setView(lng = -122.58, lat = 45.495, zoom = 11.5) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(lng = ~LONGTD_DD, lat = ~LAT_DD, 
                   data = pdx_crash_2018, radius = 1.5,
                   opacity = 0.6, group = ~day_time, 
                   color = ~pal(day_time)) %>%
  addLegend(position = "bottomright", pal = pal,
            values = pdx_crash_2018$day_time, opacity = 1,
            title = "Time of Crash")
```


e. Now add a pop-up to your interactive map that provides additional information (beyond part of the day) about the crash.  

```{r}
pdx_crash_2018 <- pdx_crash_2018 %>%
  mutate(severity = case_when(
           CRASH_SVRTY_CD == 2 ~ "fatal",
           CRASH_SVRTY_CD == 4 ~ "non-fatal injury",
           CRASH_SVRTY_CD == 5 ~ "property damage only"),
         alcohol = as.factor(ALCHL_INVLV_FLG),
         alcohol = case_when(
           alcohol == 0 ~ "no",
           alcohol == 1 ~ "yes"))

content <- paste("Time of day:", pdx_crash_2018$day_time, "</br>",
                 "Severity:", pdx_crash_2018$severity, "</br>",
                 "Alcohol involved:", pdx_crash_2018$alcohol)

leaflet(options = leafletOptions(minZoom = 11, maxZoom = 16)) %>% 
  setView(lng = -122.58, lat = 45.495, zoom = 11.5) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(lng = ~LONGTD_DD, lat = ~LAT_DD, 
                   data = pdx_crash_2018, radius = 1.5,
                   opacity = 0.6, group = ~day_time, 
                   color = ~pal(day_time), popup = content) %>%
  addLegend(position = "bottomright", pal = pal,
            values = pdx_crash_2018$day_time, opacity = 1,
            title = "Time of Crash")
```


f.  Create a leaflet graph where the user can toggle between displaying the different crash severities.

```{r}
fatal <- pdx_crash_2018 %>%
  filter(severity == "fatal")

nonfatalinj <- pdx_crash_2018 %>%
  filter(severity == "non-fatal injury")

propdmgonly <- pdx_crash_2018 %>%
  filter(severity == "property damage only")

brewer.pal(3, "Dark2")

pal2 <- colorFactor(palette = "Dark2", 
                    domain = pdx_crash_2018$severity)

leaflet(options = leafletOptions(minZoom = 11, maxZoom = 16)) %>% 
  setView(lng = -122.58, lat = 45.495, zoom = 11.5) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(lng = ~LONGTD_DD, lat = ~LAT_DD, 
                   data = fatal, radius = 4.5,
                   fillOpacity = 0.2, group = "fatal", 
                   fillColor = "#1B9E77",
                   stroke = FALSE, popup = content) %>%
  addCircleMarkers(lng = ~LONGTD_DD, lat = ~LAT_DD, 
                   data = nonfatalinj, radius = 4.5,
                   fillOpacity = 0.2, group = "non-fatal injury", 
                   fillColor = "#D95F02",
                   stroke = FALSE, popup = content) %>%
  addCircleMarkers(lng = ~LONGTD_DD, lat = ~LAT_DD, 
                   data = propdmgonly, radius = 4.5,
                   fillOpacity = 0.2, group = "property damage only", 
                   fillColor = "#7570B3",
                   stroke = FALSE, popup = content) %>%
  addLegend(position = "bottomright", pal = pal2,
            values = pdx_crash_2018$severity, opacity = 1,
            title = "Severity of crash") %>%
  addLayersControl(overlayGroups = c("fatal", "non-fatal injury",
                                     "property damage only"),
                   options = layersControlOptions(collapsed = FALSE),
                   position = "topright")
```




### Problem 2: Choropleth Maps

For this problem, I want you to practice creating choropleth maps.  Let's grab some data using `tidycensus`.  You will have to set up an [API key](https://api.census.gov/data/key_signup.html).

```{r}
api_key <- "01ce758cf0488c54a09a4c14249a4f4d604a6078"
```


a. Let's grab data on the median gross rent (`B25064_001`) from the American Community Survey for Multnomah county, Oregon.  I want you to do data pulls at three geography resolutions: `county subdivision`, `tract`, and `block group`.

```{r}
options(tigris_use_cache = TRUE)

multsub <- get_acs(geography = "county subdivision",
                   variables = "B25064_001",
                   state = "Oregon",
                   county = "Multnomah",
                   geometry = TRUE,
                   key = api_key)

multtract <- get_acs(geography = "tract",
                     variables = "B25064_001",
                     state = "Oregon",
                     county = "Multnomah",
                     geometry = TRUE,
                     key = api_key)

multblock <- get_acs(geography = "block group",
                     variables = "B25064_001",
                     state = "Oregon",
                     county = "Multnomah",
                     geometry = TRUE,
                     key = api_key)
```


b. Create three choropleth maps of gross rent, one for each geography resolution.

```{r}
multsub <- multsub %>%
  mutate(NAME = str_replace_all(NAME, " CCD, Multnomah County, Oregon", ""))

ggplot(data = multsub, mapping = aes(geometry = geometry,
                                     fill = estimate)) + 
  geom_sf(color = "white") +
  coord_sf() +
  theme_void() +
  labs(title = "Median Monthly Rent for Multnomah County, OR",
       subtitle = "by Subdivision",
       fill = "Estimate ($)", caption = "Source: 5 year ACS 2019") +
  geom_sf_label(data = multsub, inherit.aes = FALSE,
                      aes(label = NAME), size = 3) +
  scale_fill_continuous(high = "#132B43", low = "#56B1F7")
```

```{r}
ggplot(data = multtract, mapping = aes(geometry = geometry,
                                       fill = estimate)) + 
  geom_sf(color = "white") +
  coord_sf() +
  theme_void() +
  labs(title = "Median Monthly Rent for Multnomah County, OR",
       subtitle = "by Census Tract",
       fill = "Estimate ($)", caption = "Source: 5 year ACS 2019") +
  scale_fill_continuous(high = "#132B43", low = "#56B1F7")
```

```{r}
ggplot(data = multblock, mapping = aes(geometry = geometry,
                                       fill = estimate)) + 
  geom_sf(color = "white") +
  coord_sf() +
  theme_void() +
  labs(title = "Median Monthly Rent for Multnomah County, OR",
       subtitle = "by Census Block",
       fill = "Estimate ($)", caption = "Source: 5 year ACS 2019") +
  scale_fill_continuous(high = "#132B43", low = "#56B1F7")
```


c. Make one of your maps interactive.  

```{r}
pal3 <- colorNumeric(palette = "viridis", domain = multtract$estimate,
                     reverse = TRUE)

popup <- paste(multtract$NAME, "</br>",
               "Median rent ($):", multtract$estimate, "</br>",
               "Margin of error ($):", multtract$moe)

multtract %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(color = ~pal3(estimate), stroke = FALSE,
              fillOpacity = 0.85, popup = popup) %>%
  addLegend(position = "bottomright", pal = pal3,
            values = multtract$estimate, opacity = 1,
            title = "Median rent ($)")
```

Some lovely code I found [here](https://stackoverflow.com/questions/40276569/reverse-order-in-r-leaflet-continuous-legend) to fix the legend order!

```{r}
addLegend_decreasing <- function (map, position = c("topright", "bottomright", "bottomleft","topleft"),
                                  pal, values, na.label = "NA", bins = 7, colors, 
                                  opacity = 0.5, labels = NULL, labFormat = labelFormat(), 
                                  title = NULL, className = "info legend", layerId = NULL, 
                                  group = NULL, data = getMapData(map), decreasing = FALSE) {
  
  position <- match.arg(position)
  type <- "unknown"
  na.color <- NULL
  extra <- NULL
  if (!missing(pal)) {
    if (!missing(colors)) 
      stop("You must provide either 'pal' or 'colors' (not both)")
    if (missing(title) && inherits(values, "formula")) 
      title <- deparse(values[[2]])
    values <- evalFormula(values, data)
    type <- attr(pal, "colorType", exact = TRUE)
    args <- attr(pal, "colorArgs", exact = TRUE)
    na.color <- args$na.color
    if (!is.null(na.color) && col2rgb(na.color, alpha = TRUE)[[4]] == 
        0) {
      na.color <- NULL
    }
    if (type != "numeric" && !missing(bins)) 
      warning("'bins' is ignored because the palette type is not numeric")
    if (type == "numeric") {
      cuts <- if (length(bins) == 1) 
        pretty(values, bins)
      else bins   
      if (length(bins) > 2) 
        if (!all(abs(diff(bins, differences = 2)) <= 
                 sqrt(.Machine$double.eps))) 
          stop("The vector of breaks 'bins' must be equally spaced")
      n <- length(cuts)
      r <- range(values, na.rm = TRUE)
      cuts <- cuts[cuts >= r[1] & cuts <= r[2]]
      n <- length(cuts)
      p <- (cuts - r[1])/(r[2] - r[1])
      extra <- list(p_1 = p[1], p_n = p[n])
      p <- c("", paste0(100 * p, "%"), "")
      if (decreasing == TRUE){
        colors <- pal(rev(c(r[1], cuts, r[2])))
        labels <- rev(labFormat(type = "numeric", cuts))
      }else{
        colors <- pal(c(r[1], cuts, r[2]))
        labels <- rev(labFormat(type = "numeric", cuts))
      }
      colors <- paste(colors, p, sep = " ", collapse = ", ")
    }
    else if (type == "bin") {
      cuts <- args$bins
      n <- length(cuts)
      mids <- (cuts[-1] + cuts[-n])/2
      if (decreasing == TRUE){
        colors <- pal(rev(mids))
        labels <- rev(labFormat(type = "bin", cuts))
      }else{
        colors <- pal(mids)
        labels <- labFormat(type = "bin", cuts)
      }
    }
    else if (type == "quantile") {
      p <- args$probs
      n <- length(p)
      cuts <- quantile(values, probs = p, na.rm = TRUE)
      mids <- quantile(values, probs = (p[-1] + p[-n])/2, na.rm = TRUE)
      if (decreasing == TRUE){
        colors <- pal(rev(mids))
        labels <- rev(labFormat(type = "quantile", cuts, p))
      }else{
        colors <- pal(mids)
        labels <- labFormat(type = "quantile", cuts, p)
      }
    }
    else if (type == "factor") {
      v <- sort(unique(na.omit(values)))
      colors <- pal(v)
      labels <- labFormat(type = "factor", v)
      if (decreasing == TRUE){
        colors <- pal(rev(v))
        labels <- rev(labFormat(type = "factor", v))
      }else{
        colors <- pal(v)
        labels <- labFormat(type = "factor", v)
      }
    }
    else stop("Palette function not supported")
    if (!any(is.na(values))) 
      na.color <- NULL
  }
  else {
    if (length(colors) != length(labels)) 
      stop("'colors' and 'labels' must be of the same length")
  }
  legend <- list(colors = I(unname(colors)), labels = I(unname(labels)), 
                 na_color = na.color, na_label = na.label, opacity = opacity, 
                 position = position, type = type, title = title, extra = extra, 
                 layerId = layerId, className = className, group = group)
  invokeMethod(map, data, "addLegend", legend)
}

#reverse the order

multtract %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(color = ~pal3(estimate), stroke = FALSE,
              fillOpacity = 0.85, popup = popup) %>%
  addLegend_decreasing(position = "bottomright", pal = pal3,
                       values = multtract$estimate, opacity = 1,
                       title = "Median rent ($)",
                       decreasing = TRUE)
```



### Problem 4: Your Turn!


a. Using the `pdxTrees` dataset, create two `leaflet` maps.  I want you to get creative and really dig into the functionalities of `leaflet`.  Consider

* Focusing on a specific park or set of parks.
* Potentially using special icons.
* Including labels and/or pop-ups.
* The best tiling for your purpose.
* Zoom or view constraints


```{r}
#Map 1

parks <- get_pdxTrees_parks()

unique(parks$Functional_Type)

mttabor <- parks %>%
  filter(Park == "Mt Tabor Park")  %>%
  mutate(ft2 = case_when(
    Functional_Type == "BD" ~ "broadleaf deciduous",
    Functional_Type == "BE" ~ "broadleaf evergreen",
    Functional_Type == "CD" ~ "conifer deciduous",
    Functional_Type == "CE" ~ "conifer evergreen"
  ))

broadleaf <- mttabor %>%
  filter(Functional_Type %in% c("BD", "BE"))

broadleaf_icon <- makeIcon(
  iconUrl = "https://emojipedia-us.s3.dualstack.us-west-1.amazonaws.com/thumbs/120/joypixels/275/deciduous-tree_1f333.png",
  iconWidth = 10, iconHeight = 20)

conifer <- mttabor %>%
  filter(Functional_Type %in% c("CD", "CE"))

conifer_icon <- makeIcon(
  iconUrl = "https://emojipedia-us.s3.dualstack.us-west-1.amazonaws.com/thumbs/120/joypixels/275/evergreen-tree_1f332.png",
  iconWidth = 10, iconHeight = 20)

popup1 <- paste("<b>", broadleaf$Common_Name, 
                "</b></br><i>", broadleaf$Scientific_Name, 
                "</i>, family", broadleaf$Family,
                "</br>", "DBH:", broadleaf$DBH, "inches",
                "</br>", "Height:", broadleaf$Tree_Height, "feet",
                "</br>", "Annual carbon sequestration:", broadleaf$Carbon_Sequestration_lb, "pounds",
                "</br>", "Annual stormwater interception:", broadleaf$Stormwater_ft, "cubic feet",
                "</br>", "Annual pollution removal:", broadleaf$Pollution_Removal_oz, "ounces")

popup2 <- paste("<b>", conifer$Common_Name, 
                "</b></br><i>", conifer$Scientific_Name, 
                "</i>, family", conifer$Family,
                "</br>", "DBH:", conifer$DBH, "inches",
                "</br>", "Height:", conifer$Tree_Height, "feet",
                "</br>", "Annual carbon sequestration:", conifer$Carbon_Sequestration_lb, "pounds",
                "</br>", "Annual stormwater interception:", conifer$Stormwater_ft, "cubic feet",
                "</br>", "Annual pollution removal:", conifer$Pollution_Removal_oz, "ounces")

leaflet(options = leafletOptions(minZoom = 12, maxZoom = 19)) %>%
  setView(lng = -122.594848, lat = 45.511, zoom = 15) %>% 
  setMaxBounds(lat1 = 45.55, lng1 = -122.50, lat2 = 45.47, lng2 = -122.68) %>%
  addProviderTiles(providers$CartoDB.Voyager) %>%
  addMarkers(lng = ~Longitude, lat = ~Latitude,
             data = broadleaf, icon = broadleaf_icon,
             popup = popup1) %>%
  addMarkers(lng = ~Longitude, lat = ~Latitude,
             data = conifer, icon = conifer_icon,
             popup = popup2)
```


```{r}
#Map 2

near_reed <- filter(parks, Park %in% c("Crystal Springs Rhododendron Garden",
                                       "Kenilworth Park", 
                                       "Eastmoreland Garden",
                                       "Berkeley Park"))

near_reed_bl <- near_reed %>%
  filter(Functional_Type %in% c("BD", "BE"))

popup3 <- paste("<b>", near_reed_bl$Common_Name, 
                "</b></br><i>", near_reed_bl$Scientific_Name, 
                "</i>, family", near_reed_bl$Family,
                "</br>", "DBH:", near_reed_bl$DBH, "inches",
                "</br>", "Height:", near_reed_bl$Tree_Height, "feet",
                "</br>", "Annual carbon sequestration value: $", near_reed_bl$Carbon_Sequestration_value,
                "</br>", "Annual stormwater interception value: $", near_reed_bl$Stormwater_value,
                "</br>", "Annual pollution removal value: $", near_reed_bl$Pollution_Removal_value)

near_reed_con <- near_reed %>%
  filter(Functional_Type %in% c("CD", "CE"))

popup4 <- paste("<b>", near_reed_con$Common_Name, 
                "</b></br><i>", near_reed_con$Scientific_Name, 
                "</i>, family", near_reed_con$Family,
                "</br>", "DBH:", near_reed_con$DBH, "inches",
                "</br>", "Height:", near_reed_con$Tree_Height, "feet",
                "</br>", "Annual carbon sequestration value: $", near_reed_bl$Carbon_Sequestration_value,
                "</br>", "Annual stormwater interception value: $", near_reed_bl$Stormwater_value,
                "</br>", "Annual pollution removal value: $", near_reed_bl$Pollution_Removal_value)

near_reed %>%
  group_by(Park) %>%
  summarize(sum(Carbon_Sequestration_value, na.rm = TRUE),
            sum(Stormwater_value, na.rm = TRUE),
            sum(Pollution_Removal_value, na.rm = TRUE))

park_markers <- data.frame(name = c("Crystal Springs Rhododendron Garden",
                                    "Kenilworth Park", 
                                    "Eastmoreland Garden",
                                    "Berkeley Park"),
                           lat = c(45.477932, 45.491347,
                                   45.474034, 45.473294),
                           long = c(-122.637821, -122.631206,
                                    -122.636110, -122.624426),
                           carbon = c("469", "473", "6.65", "383"),
                           water = c("991", "1,270", "12.20", "756"),
                           pollution = c("1,800", "2,307", "22.20", "1,373"))

pop5 <- paste("<b>", park_markers$name,
              "</b></br>", "Park total annual carbon sequestration value: $",
              park_markers$carbon, "</br>",
              "Park total annual stormwater interception value: $",
              park_markers$water, "</br>",
              "Park total annual pollution removal value: $", 
              park_markers$pollution)

leaflet(options = leafletOptions(minZoom = 12, maxZoom = 18)) %>% 
  setView(lng = -122.63, lat = 45.4817, zoom = 14) %>% 
  setMaxBounds(lat1 = 45.5027, lng1 = -122.58, lat2 = 45.4607, lng2 = -122.68) %>%
  addProviderTiles(providers$Esri.WorldStreetMap) %>%
  addMarkers(lng = ~Longitude, lat = ~Latitude, 
             data = near_reed_bl, icon = broadleaf_icon,
             popup = popup3) %>%
  addMarkers(lng = ~Longitude, lat = ~Latitude, 
             data = near_reed_con, icon = conifer_icon,
             popup = popup4) %>%
  addMarkers(lng = ~long, lat = ~lat,
             data = park_markers, popup = pop5)
```
